Live Functional Programming with Typed Holes

### 2018-04-10 appendix-defns.tex
finished draft of 3.2
finished draft of sec3 and 4

@@

>>>>

\subsection{Substitution}
\label{sec:substitution}
\judgbox
  {[\dexp/x]\dexp' = \dexp''}
  {$\dexp''$ is obtained by substituting $\dexp$ for $x$ in $\dexp'$}
  %% {$\dexp''$ is the result of substituting $\dexp$ for $u$ in $\dexp'$}

\vspace{5pt}
\judgbox
  {[\dexp/x]\sigma = \sigma'}
  {$\sigma'$ is obtained by substituting $\dexp$ for $x$ in $\sigma$}
  %% {$\dexp''$ is the result of substituting $\dexp$ for $u$ in $\dexp'$}
\[
\begin{array}{lcll}
\substitute{\dexp}{x}{c}
&=&
c\\
\substitute{\dexp}{x}{x}
&=&
\dexp
\\%
\substitute{\dexp}{x}{y}
&=&
y
& \text{when $x \neq y$}
\\
\substitute{\dexp}{x}{\halam{y}{\htau}{\dexp'}}
&=&
\halam{y}{\htau}{\substitute{\dexp}{x}{\dexp'}}
& \text{when $x \neq y$ and $y \notin \fvof{d}$}
\\
\substitute{\dexp}{x}{\dap{\dexp_1}{\dexp_2}}
&=&
\dap{(\substitute{\dexp}{x}{\dexp_1})}{\substitute{\dexp}{x}{\dexp_2}}
\\
\substitute{\dexp}{x}{\dehole{u}{\subst}{}}
&=&
\dehole{u}{\substitute{\dexp}{x}{\subst}}{}
\\
\substitute{\dexp}{x}{\dhole{\dexp'}{u}{\subst}{}}
&=&
\dhole{\substitute{\dexp}{x}{\dexp'}}{u}{\substitute{\dexp}{x}{\subst}}{}
\\
\substitute{\dexp}{x}{\dcasttwo{\dexp'}{\htau_1}{\htau_2}}
&=&
\dcasttwo{(\substitute{\dexp}{x}{\dexp'})}{\htau_1}{\htau_2}
\\
\substitute{\dexp}{x}{\dcastfail{\dexp'}{\htau_1}{\htau_2}}
&=&
\dcastfail{(\substitute{\dexp}{x}{\dexp'})}{\htau_1}{\htau_2}
\\
\substitute{\dexp}{x}{\cdot}
&=&
\cdot\\
\substitute{\dexp}{x}{\sigma, d/y}
&=&
\substitute{\dexp}{x}{\sigma}, \substitute{\dexp}{x}{d}/y
%% {[\dexp_1 / x] \dcast{\htau}{\dexp}}
%% &=&
%% \dcast{\htau}{[\dexp_1 / x] \dexp}
\end{array}
\]

\vspace{5pt}
\begin{lem}[Substitution] \label{thm:substitution}~
\begin{enumerate}[nolistsep]
\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{d}{\htau}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]d}{\htau}$.
\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{\sigma}{\hGamma'}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]\sigma}{\hGamma'}$.
\end{enumerate}
\end{lem}

\subsection{Canonical Forms} 
\begin{lem}[Canonical Value Forms]\label{thm:canonincal-value-forms}
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isValue{\dexp}$
  then $\htau\neq\tehole$ and
  \begin{enumerate}[nolistsep]
    \item If $\htau=b$ then $\dexp=c$.
    \item If $\htau=\tarr{\htau_1}{\htau_2}$
          then $\dexp=\halam{x}{\htau_1}{\dexp'}$
          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$.
  \end{enumerate}
\end{lem}

\begin{lem}[Canonical Boxed Forms]\label{thm:canonical-boxed-forms}
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isBoxedValue{\dexp}$
  then
  \begin{enumerate}[nolistsep]
    \item If $\htau=b$ then $\dexp=c$.
    \item If $\htau=\tarr{\htau_1}{\htau_2}$ then either
      \begin{enumerate}
        \item[i.]
          $\dexp=\halam{x}{\htau_1}{\dexp'}$
          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$, or
        \item[ii.]
          $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$
          where $\tarr{\htau_1'}{\htau_2'}\neq\tarr{\htau_1}{\htau_2}$
          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\tarr{\htau_1'}{\htau_2'}}$.
      \end{enumerate}
    \item If $\htau=\tehole$
          then $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
          where $\isGround{\htau'}$
          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$.
  \end{enumerate}
\end{lem}

\begin{lem}[Canonical Indeterminate Forms]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isIndet{\dexp}$
  then either\todo{expand this to have one case per type}
  \begin{enumerate}[nolistsep]
    \item
      $\dexp=\dehole{u}{\subst}{}$
      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
    \item
      $\dexp=\dhole{\dexp'}{u}{\subst}{}$
      and $\isFinal{\dexp'}$
      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$
      and $\Dbinding{u}{\Gamma'}{\htau}\in\hDelta$, or
    \item
      $\dexp=\dap{\dexp_1}{\dexp_2}$
      and $\hasType{\hDelta}{\emptyset}{\dexp_1}{\tarr{\htau_2}{\htau}}$
      and $\hasType{\hDelta}{\emptyset}{\dexp_2}{\htau_2}$
      and $\isIndet{\dexp_1}$
      and $\isFinal{\dexp_2}$
      and $\dexp_1\neq\dcasttwo{\dexp_1}{\tarr{\htau_3}{\htau_4}}
                                        {\tarr{\htau_3'}{\htau_4'}}$, or
    %% \item
    %%   \begin{enumerate}
    %%     \item blah
    %%     \item blah
    %%     \item blah
    %%   \end{enumerate}
    \item
      $\htau=b$
      and $\dexp=\dcasttwo{\dexp'}{\tehole}{b}$
      and $\isIndet{\dexp'}$
      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
    \item
      $\htau=b$
      and $\dexp=\dcastfail{\dexp'}{\htau'}{b}$
      and $\isGround{\htau'}$
      and $\htau'\neq{b}$
      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
    \item
      $\htau=\tarr{\htau_{11}}{\htau_{12}}$
      and $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1}{\htau_2}}
                                  {\tarr{\htau_{11}}{\htau_{12}}}$
      and $\isIndet{\dexp'}$
      and $\tarr{\htau_1}{\htau_2}\neq\tarr{\htau_{11}}{\htau_{12}}$, or
    \item
      $\htau=\tarr{\tehole}{\tehole}$
      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
      %% and $\htau_{11}=\tehole$
      %% and $\htau_{12}=\tehole$
      and $\dexp=\dcastthree{\dexp'}{\tehole}{\tehole}{\tehole}$
      and $\isIndet{\dexp'}$
      and $\dexp'\neq\dcasttwo{\dexp''}{\htau'}{\tehole}$, or
    \item
      $\htau=\tarr{\tehole}{\tehole}$
      %% $\htau=\tarr{\htau_{11}}{\htau_{12}}$
      %% and $\htau_{11}=\tehole$
      %% and $\htau_{12}=\tehole$
      and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\tehole}{\tehole}}$
      %% and $\dexp=\dcastfail{\dexp'}{\htau'}{\tarr{\htau_{11}}{\htau_{12}}}$
      and $\htau'\neq\htau$
      and $\isGround{\htau'}$
      and $\isIndet{\dexp'}$
      and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$, or
    \item
      $\htau=\tehole$
      and $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
      and $\isGround{\htau'}$
      and $\isIndet{\dexp'}$.
  \end{enumerate}
\end{lem}

% No weakening for Gammas in Delta: 
% If $\hasType{\Delta, \Dbinding{u}{\hGamma}{\tau}}{\hGamma'}{d}{\tau'}$ then $\hasType{\Delta, \Dbinding{u}{\hGamma, x : \tau''}{\tau}}{\hGamma'}{d}{\tau'}$.

\clearpage
\subsection{Complete Programs}

@@

\clearpage
\subsection{Hole Filling}\label{sec:hole-filling}
\input{fig-multi-step}
\input{fig-evalctx-instantiation}
\begin{lem}[Filling] ~
  \begin{enumerate}[nolistsep]
  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\dexp}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\sigma}{\hGamma''}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\sigma}}{\hGamma''}$.
  \end{enumerate}
\end{lem}
\begin{proof}
In each case, we proceed by rule induction on the first assumption, appealing to the Substitution Lemma as necessary.
\end{proof}

\begin{lem}[Substitution Commutativity]
  If 
  \begin{enumerate}[nolistsep]
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{x : \htau_2}{\dexp_1}{\tau}$ and 
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_2}{\htau_2}$ and 
  	\item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  \end{enumerate}

  then  $\instantiate{d'}{u}{\substitute{d_2}{x}{d_1}} = \substitute{\instantiate{d'}{u}{d_2}}{x}{\instantiate{d'}{u}{d_1}}$.
\end{lem}

\begin{lem}[Instruction Commutativity]
  If 
  \begin{enumerate}[nolistsep]
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$ and 
  	\item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and 
  	\item $\reducesE{}{\dexp_1}{\dexp_2}$
  \end{enumerate}

  then $\reducesE{}{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{lem}

\begin{lem}[Filling Totality]
Either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx}=\evalctx'$ for some $\evalctx'$.
\end{lem}

\begin{lem}[Discarding] If 
	\begin{enumerate}[nolistsep]
	\item $\selectEvalCtx{d_1}{\evalctx}{\dexp_1'}$ and 
	\item $\selectEvalCtx{d_2}{\evalctx}{\dexp_2'}$ and 
	\item $\inhole{u}{\evalctx}$
	\end{enumerate}

	then $\instantiate{d}{u}{d_1} = \instantiate{d}{u}{d_2}$.
\end{lem}

\begin{lem}[Filling Distribution] If 
	$\selectEvalCtx{d_1}{\evalctx}{d_1'}$ and $\instantiate{d}{u}{\evalctx}=\evalctx'$ then $\selectEvalCtx{\instantiate{d}{u}{d_1}}{\evalctx'}{\instantiate{d}{u}{d_1'}}$.
\end{lem}

\begin{thm}[Commutativity]
  If 
  \begin{enumerate}[nolistsep]
  \item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$ and 
  \item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and 
  \item $\multiStepsTo{\dexp_1}{\dexp_2}$
\end{enumerate}

  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
\begin{proof}
By rule induction on (3). The reflexive case is immediate. In the inductive case, we proceed by rule induction on the stepping premise. There is one rule, Rule~{Step}. By Filling Totality, either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx} = \evalctx'$. In the former case, by Discarding, we can conclude by MultiStepRefl. In the latter case, by Instruction Commutativity and Filling Distribution we can take a step, and we can conclude via MultiStepSteps by applying Preservation and then the IH.
\end{proof}
